---
title: "MIKE analysis for Africa - Summary"
author: "Carl Schwarz"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  pdf_document:
    number_sections: yes
    toc: yes
    fig_caption: true
    extra_dependencies: ["float"]
  html_document:
    number_sections: yes
    toc: yes
---

```{r setup, echo=FALSE,message=FALSE, warning=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE)
options(width=200)

library(ggplot2)
library(ggforce)
library(ggmap)
library(gridExtra)
library(HDInterval)
library(kableExtra)
library(knitr)
library(plyr)
library(reshape2)
library(R2jags)
library(rgdal)
library(tidyverse)
library(tidyxl)
library(viridis)

source("register_google_key.R")


# load the fitting functions
source("fitting-functions1-base-model.R")  # non-hierarchical model for year effect
source("fitting-functions1-base-model-noSY.R")  # non-hierarchical model for year effect
source("fitting-functions2.R")  # extraction and other functions


source("read.africa.data.R", echo=TRUE)



```

# Introduction

In this document, we summarize how yearly-trends in the Proportion of Illegally Killed Elephants (PIKE) from 
MIKE (Monitoring Illegally Killed Elephants) monitoring sites are found. These are first found at the site level
and then aggregated to the continental and sub-regional level.

## Survey protocol

Briefly, MIKE data is collected on an annual basis in designated MIKE sites by law enforcement 
and ranger patrols in the field and through other means. 
When an elephant carcass is found, site personnel try to establish the cause of death and other details, 
such as sex and age of the animal, status of ivory, and stage of decomposition of the carcass. 
This information is recorded in standardized carcass forms, details of which are then submitted to the 
MIKE Programme. 
As expected, different sites report widely different numbers of carcasses, 
as encountered carcass numbers are a function of: population abundance; natural mortality rates; 
the detection probabilities of elephant carcasses in different habitats; 
differential carcass decay rates; 
levels of illegal killing; and levels of search effort and site coverage. 
Because of these features of the survey data, the number of carcasses found is unlikely to be 
proportional to the total mortality and trends in observed numbers of 
illegally killed elephants may not be informative of the underlying trends in poaching pressure. 

## Definition of *PIKE*

Consequently, the observed proportion of illegally killed elephants (PIKE) as an index of 
poaching levels has been used in the MIKE analysis in an attempt to account for 
differences in patrol effort between sites and over time:
$$PIKE_{sy}=\frac{\textit{Number of illegally killed elephants}_{sy}}{\textit{Total Carcasses Examined}_{sy}}$$
where the subscripts $sy$ refer to site and year respectively.

Computing a continent-wide PIKE is challenging for several reasons, including as mentioned above:

- Detection probabilities of elephant carcasses in various habitats differ.
- Levels of search effort and site coverage differ between sites.
- Not all sites report in all years.
- Number of carcasses in both categories varies considerably across space and time.

In past years, a simple linear model on the PIKE values was computed 
as described in https://cites.org/sites/default/files/notif/E-Notif-2019-046.pdf.
This is denoted the *LSMeans* approach.

The PIKE trend is calculated using estimated marginal means of a linear model weighted by 
the total number of observed carcasses. 
The continental PIKE trend is estimated based on a model with 
subregion and year as factors, while the subregional trends are estimated from a 
model using country and year as factors.

As indicated in the MIKE report to the 18th meeting of the Conference of Parties to CITES available at
https://cites.org/sites/default/files/eng/cop/18/doc/E-CoP18-069-02.pdf,
the CITES Secretariat, in collaboration with the MIKE-ETIS TAG statisticians and an independent statistician, 
initiated a process to review the MIKE analytical methodology to determine 
whether it could be refined or its scientific robustness improved, 
and to further enhance the analytical basis for MIKE. 
The approach included a review of the current methodology, and consideration of new statistical developments and, 
therefore, alternative methods or models for *PIKE* trend analysis, 
while taking into consideration the imbalances and inconsistencies inherent in the data. 
Burn, Underwood and Blanc (2011) used
a Bayesian hierarchical model based on a generalized linear mixed model (logistic regression with random effects) to resolve
many of these issues.

Issues with the collection of the data such as:

- search patterns are not random and often directed to illegally killed elephants;
- misclassification of animals as unknown causes of mortality rather than illegally or natural mortality
based on the preponderance of evidence;
- errors in the data

are not considered here, but the CITES MIKE CCU is preparing discussion papers that explains these issues in more detail.

This document is a brief summary of the result of using a GLMM to estimate trends in *PIKE*. A technical
document is available with detailed results, model assessment, and a sensitivity analysis.


## Why change from the LSMeans approach

The *LSMeans* approach has been used for the *PIKE* trend analysis in the reports to
previous meetings of the Conference of the Parties (CoP16, Bangkok, 2013 in document CoP16 Doc. 53.1; and 
CoP17, Johannesburg, 2016, in document CoP17 Doc. 57.5); 
and to meetings of the Standing Committee 
(SC62, Geneva, July 2012, in document SC62 Doc. 46.1 (Rev. 1); 
SC65, Geneva, July 2014, in document SC65 Doc. 42.1;
SC66, Geneva, January 2016, in document SC66 Doc. 47.1; 
SC69, Geneva, November 2017, in document SC69 Doc. 51.1;
and SC70, Sochi, October 2018, in document SC70 Doc. 49.1).
However, there are a number of issues with this approach

- Predicted *PIKE* values could be less than 0 or greater than 1
- Each country is given equal weight regardless of the number of *MIKE* sites or the abundance of elephants
in each site at the sub-regional level. For example, in Eastern Africa, both Kenya (4 *MIKE* sites) and Eritrea (1 *MIKE* site) are given equal weight in computing
the subregional PIKE. But at the continental level, countries are ignored and data pooled over all sites in a sub-region.
- The *LSMeans* is not consistent in how to aggregate to larger levels.
The mean of the *LSMeans* estimates of *PIKE* at the sub-regional level from the individual sub-regional models, will not match the estimated
*PIKE* at the continental level estimated using the continental model because country is or is not included in the two models.
- It is difficult to apply different weighting, e.g. by elephant populations at the *MIKE* or country level.
- The *LSMeans* approach imputes *PIKE* for missing country-year combinations at the sub-regional level, 
but the individual *MIKE* sites could differ in the patterns of missingness across years. Consequently, 
the country may appear to have data for every year, but the country-wide aggregate data is based on a different
combination of *MIKE* sites across the years. A similar problem occurs in the analysis at the continental level.
- Not all sources of variation are included in estimates of uncertainty. 
For example, binomial variation (Section 5.2 of the technical document) at the
site-year level is not included, i.e. if the actual *PIKE* of all elephants at the *MIKE* site is .20, then the observed
*PIKE* in the sample of carcasses examined will vary around 0.20.

The impact of giving equal weight at the country level or at the *MIKE* site level is explored in more detail in Appendix 3
where this has a noticeable impact on the sub-regional *PIKE* for East Africa.
 
 
## Advantages of GLMM

At the recent MIKE-ETIS TAG meeting (September 2019, Nairobi), the use of the generalized linear mixed model (GLMM) was recommended
going forward. This document provides an analysis of the *PIKE* data using a GLMM, compares the results to 
those from the previous analyses, and explores the impacts of various assumptions on the estimated *PIKE*.

The advantages of the GLMM approach are:

- The new model fully accounts for the binomial structure at the site-year level, i.e. of $n$ carcasses observed, $x$
are illegally killed.
- The new model fully accounts for different sample sizes, i.e. a *PIKE* based on observing 1 illegally killed
elephants out of 2 elephant carcasses is given different weight than a *PIKE* based on observing 20 illegally killed
elephants out of 40 elephant carcasses.
- Imputation of missing data takes place at the MIKE site level based on the relationship of *PIKE* in this site
to the *PIKE* at other sites across time.
- Each *MIKE* site is given equal weight when computing the continental or sub-regional *PIKE*. Consequently,
countries with more *MIKE* sites will automatically be given more weight in the aggregate *PIKE* estimates. 
- It is easy to apply other weightings, e.g. by the elephant population abundances at each *MIKE* site when computing
an aggregate *PIKE*
- Multiple sources of variation are automatically included, e.g. binomial variation at the site-year level,
site-year interactions representing the changes in *PIKE* over time at a particular *MIKE* site, and
site-to-site variation. This variation is automatically included in the uncertainty of the aggregate *PIKE* estimates.
- Uncertainty of the estimated *PIKE* at the continental or sub-regional level can be computed assuming that the
current *MIKE* sites are index sites (and fixed) or random sample of potential *MIKE* sites. The current design
is somewhat between these two extremes, and so the uncertainty reported under these two ways of viewing the
current set of *MIKE* sites represents a lower and upper bound of uncertainty.
- It is easy to include estimates of uncertainty in elephant abundances when weighting *MIKE* sites by 
elephant abundances. This is currently not yet done because of issues in determining the uncertainty in these
abundance estimates (see Section 8.7 of the technical document).
- The current model implicitly accounts for spatial autocorrelation in the *PIKE* among *MIKE* sites that are
geographically close. The estimated site-level effects are similar for sites that are geographically close and 
have similar levels of governance and poaching.
- It will be possible to extend this model to account explicitly for spatial and spatial-temporal 
autocorrelation (Zuur, 2019 )

## *PIKE* as an index of poaching pressure

The value of *PIKE* computed by *LSMeans* and the *GLMM* approach should be considered as an INDEX of poaching pressure. 
We hope that trends in the index reflect trends in the actual levels of poaching. 
Converting the value of *PIKE* into a measure of actual poaching mortality is complicated due to the following:

- Effects of changing natural mortality over time are confounded with changes in *PIKE*.
- MIKE sites are not selected at random and so the marginal mean *PIKE* may not be representative 
of the actual marginal mean *PIKE*.
- Poaching at MIKE sites may not be representative of poaching in areas outside of the MIKE site.
- The unweighted marginal mean *PIKE* gives equal weight to every MIKE site. We have investigated
in the technical report the impact of weighting sites by the population abundance of elephants associated with
that site, but have not included estimates of uncertainty in the population abundance estimates.
- There are a number of issues with the collection of the data such as:
   + Management related deaths are included in the number of carcasses examined (the denominator of *PIKE*)
at this point in time;
   + search patterns are not random and often directed to illegally killed elephants;
   + misclassification of animals as unknown causes of mortality rather than illegally or natural mortality
based on the preponderance of evidence;
   + errors in the data
   
The CITES MIKE CCU is preparing a discussion paper that explores data issues in more detail.

For these reasons, great care should to be taken and assumptions should be well documented 
when converting the estimated *PIKE* to actual levels of poaching mortality.




# Exploration of PIKE data

## Location of MIKE sites
There are `r sum(data.source$GIS)` 
MIKE sites in Africa broken into 4 regions. 
Data from only `r length(unique(pike$MIKEsiteID))` MIKE sites across the sub-regions are used in the analysis because
some sites have never submitted data or are currently lacking population abundance data. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
mike.location.plot <- base.map +
   ggtitle(paste(UNRegion.select,": Location of MIKE sites",sep=""))+
   geom_point(data=mike.centers[mike.centers$UNRegion==UNRegion.select,], 
              aes(y=lat, x=lon, shape=id_subreg, color=id_subreg), size=2 )+
   ylab("Latitude")+xlab("Longitude")+
    theme(legend.position=c(0,0), legend.justification=c(0,0), legend.background=element_rect(fill="transparent"))+
   scale_color_discrete(name="Subregion")+
   scale_shape_discrete(name="Subregion")
mike.location.plot
```

## MIKE sites with *PIKE* data

In addition to the above, some sites have reported 0 (zero) carcasses detected in some years.

The current analysis treats site that did not report on any carcasses in a year (no patrol effort) and
a site that reports 0 carcasses examined in year (patrol effort but no carcasses found) in the same way. This is 
because information on patrol effort is not currently used in the analysis and only the number of carcasses examined
and the number of illegally killed elephants in the sample of carcasses is used. In the latter case, 0 illegal
carcasses out of 0 carcasses examined gives a *PIKE* for that site-year of 0/0 which is indeterminate and cannot be used
in any mathematical analysis of *PIKE*.

The following plot shows that there are some sites that have reported data 
for at least one carcass in as little as one year, but other sites
have reported data for at least one carcass in almost every year.

```{r echo=FALSE,message=FALSE,warning=FALSE}
# Create a plot of which site measured in which year
# Create a plot of which site measured in which year
temp <- pike.original
temp$Carcass.cat <- car::recode(temp$TotalNumberOfCarcasses,
                            " 1:3 = '01-03';
                              3:10= '03-10';
                             11:20= '11-20';
                             21:50= '21-50';
                             51:hi= '51+';
                            ")

site.plot <- ggplot(data=temp, aes(x=year, y=MIKEsiteName, color=Carcass.cat))+
  ggtitle(paste(UNRegion.select, ": When is each site measured",sep=""))+
  geom_point()+
  ylab("Site")+xlab("Year")+
  facet_wrap(~SubregionName, ncol=2, nrow=2, scales='free_y')+
  scale_color_viridis(discrete=TRUE, name='Number\nof\ncarcasses', direction=-1 )
site.plot
```

In total, there are `r nrow(unique(pike[,c("year","MIKEsiteID")]))` unique site-years in `r UNRegion.select` 
since `r min(pike$year)` where data has been reported
(and the number of reported carcasses > 0).

The number of carcasses reported in each site-year since `r min(pike$year)` varies enormously from
`r min(pike$TotalNumberOfCarcasses)` to `r max(pike$TotalNumberOfCarcasses)` carcasses.

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data=pike, aes(x=as.factor(year), y=TotalNumberOfCarcasses))+
  ggtitle(paste(UNRegion.select,": Carcasses observed",sep=""))+
  geom_point(position=position_jitter(w=0.2))+
  geom_boxplot( alpha=0.2, outlier.size=0)+
  facet_wrap(~SubregionName, ncol=2,scale="free")+
  scale_x_discrete(breaks=seq(2000,2020,5))+
  ylab("Total number of carcasses")+xlab("Year")
```

The unusual data point for West Africa where one site reported a large number of carcasses in one year is correct and corresponds 
to the MIKE site Gourma (GOU) with total number of carcasses equal 
to 134, of which  130  were poached by armed groups who entered the area.

## Observed *PIKE*

A plot of the observed *PIKE* values from each site-year shows a wide range in the observed *PIKE* values, 
but many of the observed *PIKE* values close to 0 or 1 occur in
sites with only a small number of carcasses examined in a year: 

```{r echo=FALSE, message=FALSE, warning=FALSE}
pike.boxplot <- ggplot(data=pike, aes(x=as.factor(year), 
                                      y=NumberOfIllegalCarcasses/TotalNumberOfCarcasses) )+
  ggtitle(paste(UNRegion.select,": Observed PIKE values",sep=""),
          subtitle="Points jittered to prevent overplotting")+
  #geom_point( position=position_jitter(h=0.2, w=0.2), size=TotalNumberOfCarcasses))+
  geom_point( aes(size=TotalNumberOfCarcasses), position=position_jitter(w=0.2))+
   geom_boxplot( alpha=0.2, outlier.size=0)+
  facet_wrap(~SubregionName, ncol=2)+
  scale_x_discrete(breaks=seq(2000,2050,5))+
  scale_size_continuous(name="Total\ncarcasses\nin a\nyear")+
  xlab("Year")+ylab("Observed Pike")
pike.boxplot
```

The trend in the observed *PIKE* values for each site is:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# compute the total number of carcasses reported for each site.
total.carcass <- plyr::ddply(pike,"MIKEsiteID", plyr::summarize, 
                             total.c = sum(TotalNumberOfCarcasses))
plotdata <- merge(pike, total.carcass)

plotlist <- plyr::llply(unique(pike$SubregionName), function(SubregionName, pike){
  # select the data for this subregion
  #browser()
  pike <- pike[ pike$SubregionName == SubregionName,]
  pike$pike <- pike$NumberOfIllegalCarcasses/pike$TotalNumberOfCarcasses
  myplot <- ggplot(data=pike, aes(x=year, y=pike, color=MIKEsiteID))+
   #ggtitle(paste(UNRegion.select,": Observed pike values for each site"),
  ##        subtitle="Thicker/darker lines represent sites with more carcasses reported")+
   geom_line(alpha= .2+ pike$total.c/max(pike$total.c)*.8,
             size = .2+ pike$total.c/max(pike$total.c)*1.3 )+
   facet_wrap(~SubregionName, ncol=1,nrow=1)+
   scale_color_discrete(name="MIKE\nsite\nID")+
   xlab("Year")+ylab("Observed PIKE")+ylim(0,1)+
   guides(color=FALSE)
  myplot
}, pike=plotdata)

plotset <- arrangeGrob(grobs=plotlist, nrow=2, ncol=2, widths=c(3,3),  heights=c(3,3),
                       top=paste(UNRegion.select,": Observed pike values for each site \n Thicker/darker lines represent more carcasses"))
plot(plotset)

```

Note that with a small number of carcasses reported (e.g. 0 or 1) it is quite common 
for the reported *PIKE* to be 0 or 1 because either none or all of the carcasses have 
been illegally killed. Consequently, the trends are difficult to interpret for many sites with
only a few carcasses reported.

Patrol effort varies considerably among sites, so the number of examined carcasses only bears
a weak relationship to estimated population abundances around the MIKE sites:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Look at relationship of number of carcasses to estimated population abundances
#any(duplicated( pike[, c("MIKEsiteID","year")]))
#mike.pop.est[duplicated( mike.pop.est[, c("MIKEsiteID","year")]),]

plotdata <- merge(pike, mike.pop.est, all.x=TRUE)
ggplot(data=plotdata, aes(x=population, y=TotalNumberOfCarcasses))+
   ggtitle(paste(UNRegion.select,": Relationship between number of reported carcasses \nand est pop abundance"))+
   geom_point()+
   geom_smooth(se=FALSE)+
   xlab("Estimated population abundance")+
   ylab("Total number of carcasses examined")+
   facet_wrap(~SubregionName, ncol=2, scale="free")

```

Consequently, the total number of carcasses found is unlikely to be a consistent fraction of the
population abundance and the number of illegally killed elephant carcasses is not a good index 
for poaching pressure. Because the *PIKE* index does not depend on effort, it is thought to be 
a better indicator of poaching pressure than the observed number of carcasses of illegally killed elephants.


## MIKE sites with population data

Population data is available for `r sum(data.source$Population)` sites and has
been extracted from Thouless et al (2016).
Population surveys are not done each year and population numbers are not updated
between population surveys. 
In these cases, for purposes of illustration, these missing values were imputed 
using the most recent abundance value.
For example, if the estimated population abundances from a survey
conducted in 2010 was 500 elephants and from a survey conducted in 2015 was 400 elephants, then
the imputed population abundances for 2011, 2012, 2013, and 2014 is also 500 elephants.

For some sites, no population surveys have been done, and population estimates are "best guesses".

These population values are NOT the number of elephants in the MIKE site. Rather, the MIKE sites were
chosen to be representative of a broader population and the population values here refer to this
broader population.
**The use of population values to compute a population-weighted PIKE is currently
experimental and being evaluated and the actual implementation may change in the future.**


A plot of when an estimate of the population abundance is available  by MIKE site is:

```{r echo=FALSE,message=FALSE,warning=FALSE}
# Create a plot of which site measured in which year
mike.pop.est.original$pop.cat <- car::recode(mike.pop.est.original$population,
                                      "  0        = '0';
                                        1:50      ='1 - 50';
                                       50:200     ='50 - 200';
                                      200:999     ='200 - 999';
                                      1000:10000  ='1000 - 10,000';
                                      10000:50000 ='10,000 - 50,000';
                                      50000:hi    ='50,000+';
                                      ")
mike.pop.est.original$pop.cat <- factor(mike.pop.est.original$pop.cat,
                       levels=c("0","1 - 50","50 - 200","200 - 999","1000 - 10,000","10,000 - 50,000","50,000+"), order=TRUE)
#xtabs(~pop.cat, data=mike.pop.est.original, exclude=NULL, na.action=na.pass)

pop.plot <- ggplot(data=mike.pop.est.original, aes(x=year, y=MIKEsiteName, color=pop.cat))+
  ggtitle(paste(UNRegion.select, ": Population values available for MIKE sites",sep=""),
          subtitle="X's indicate when a population survey conducted")+
  geom_point()+
  geom_text(data=mike.pop.survey.years, aes(color=NULL), label="X", color="red", alpha=0.7)+
  ylab("Site")+xlab("Year")+
  facet_wrap(~SubregionName, ncol=2, nrow=3, scales='free_y')+
  scale_color_viridis(discrete=TRUE, name='Population', direction=-1 )
pop.plot
```

## Final dataset used

The final dataset is an amalgamation of the MIKE population data and observed *PIKE* data. Only those sites
for which population data is available and for which at least one carcass has been reported over the period
of interest are used. 

The following table identifies the MIKE sites not included in the analysis 
based on data from one of the data sources being absent:

```{r echo=FALSE, message=FALSE, warning=FALSE}
select <- apply(!data.source[,-1],1,any)
temp <- data.source[select,]
temp$MIKEsiteID <- as.character(temp$MIKEsiteID)
temp <- temp[ order(temp$MIKEsiteID),]
kable(temp[,c("MIKEsiteID","Population","pike","GIS")], row.names=FALSE,
      caption="Summary of MIKE sites where data not present in all sources",
      col.names=c("MIKEsiteID","Population values available","PIKE reported","MIKE Center Available")) %>%
      column_spec(column=c(1),         width="3cm") %>%
      column_spec(column=c(2,3,4),     width="1.5cm") %>%
      kable_styling("bordered",position = "center", full_width=FALSE, latex_options = "HOLD_position")

```


The final data set consists of `r length(unique(pike$MIKEsiteID))` MIKE sites from
`r min(pike$year)` to `r max(pike$year)` over the subregions as shown below:

```{r echo=FALSE, message=FALSE, warning=FALSE}
temp <- plyr::ddply(pike,"SubregionName", function(x){
   Site.Year =nrow(x)
   n.sites   =length(unique(x$MIKEsiteID))
   u.sites   = sort(unique(x$MIKEsiteID))
   mean.carcass= round(mean(x$TotalNumberOfCarcasses, na.rm=TRUE),1)
   sites     = paste(u.sites, collapse=", ")
   data.frame(n.sites, Site.Year,  mean.carcass,sites)
})
kable(temp, 
      caption="Summary of MIKE sites used in analysis",
      col.names=c("Subregion Name","Number of sites","# Site-Years ","Mean # carcasses reported per year","Site IDs"),
      digits=c(NA,0,0,1,NA))  %>% 
      column_spec(column=c(1),         width="3cm") %>%
      column_spec(column=c(2,3,4),     width="2cm") %>%
      column_spec(column=5,            width="5cm") %>%
      kable_styling("bordered",position = "center", full_width=FALSE, latex_options = "HOLD_position")
```


```{r echo=FALSE, warning=FALSE, messge=FALSE}
# cleanup
rm(plotdata, pike.boxplot, site.plot, mike.location.plot)
###############################################################################################3
##############################################################################################3
#############################################################################################3
```


# Generalized Linear Mixed Model (GLMM)

## Model description

The Generalized Linear Mixed Model (GLMM) is a generalization of the simple regression model currently used
(the *LSMeans* approach). The GLMM begins by modeling the multiple sources of variation present
in the data

* Binomial variation in the number of illegally killed elephants in the number of carcasses detected. For
example, the true underlying *PIKE* at a MIKE site may be 30%. However, the rangers will not detect all carcasses at the
MIKE site. Suppose that 10 carcasses are detected. Statistical theory says that if the underlying *PIKE* is 30%, then
the number of illegally killed elephants is likely to range from 1 to 6 in the sample of 10 selected due to random chance.

* Year-to-year changes in the continental *PIKE* due to factors that change from year-to-year. These are called the
**year effects**.

* Some MIKE sites tend to have higher than average *PIKE* year-after-year. These **site-effects** could be due to
factors that are consistent for a site across years.

* Within a MIKE site, the *PIKE* will vary from year-to-year within the MIKE site due to factors that vary locally
from year to year. These are known as **site-year** effects. 

The model will estimate the year, site, and site-year effects,
and then use them to impute the *PIKE* in years where a MIKE site has not reported any
carcass returns. This imputation is based on the year trends in other sites; the relation of *PIKE* among sites (i.e. some sites
tend to always have higher *PIKE* values); 
and a range in site-year effect based on what is observed in sites with data. This imputation
step then gives a *PIKE* value for every site in every year. A range of values is generated for each imputed-value to account
for the uncertainty in each of the individual effects.

Once the *PIKE* is found for every site in every year 
(some site-years are observed; site-years with missing data have *PIKE* imputed), the
marginal mean *PIKE* is found in two ways:

* Unweighted marginal mean *PIKE* as the simple average of the *PIKE* in a year over all sites.
* Weighted marginal mean *PIKE* where the weights are the estimated population abundance
of elephants represented by the MIKE site in each year.

The population abundances are allowed to vary within a MIKE site over time. 
Uncertainty in the estimated population abundances is not incorporated
into this analysis.

**The use of population values to compute a population-weighted PIKE is currently
experimental and being evaluated and the actual implementation may change in the future.**

## Model fitting

There are many ways to fit a model to data. Simple regression models use least squares where the sum of the squared
vertical deviations between the fitted line and the observed data is minimized. 
The least squares approach is a special case of a more general approach called 
maximum likelihood. However, maximum likelihood methods often perform poorly (e.g. failure to converge)
with models with multiple levels
(as in our model)
and it is difficult to create custom summaries (such as marginal means) using likelihood methods.

For this reason, we have implemented our model in a Bayesian context using a method called 
Markov Chain Monte Carlo (MCMC) sampling
that deals with many of the technical difficulties found in a likelihood fit without a great deal of effort.
Details are available in the technical document.

A Bayesian model combines information from the prior beliefs about the values of certain parameters
and the information about these parameters from the data (through a likelihood function). 
If there is a large amount of data, the information
about the parameters would usually overwhelm the information from the prior beliefs, but in cases
of sparse data, the prior beliefs may be more important. The end product of a Bayesian analysis is the
posterior distribution of belief about a parameter.
In the context of this *PIKE* analysis, we have used
vague priors with little information about the parameters so the final result is driven almost entirely by
the data from the MIKE program.

The concept of posterior belief, while technically challenging to compute, has an intuitive 
understanding that many people practice. For example, suppose you are waiting to be picked up by a friend.
But the friend is late. *A priori* (prior belief), you could place different weights on two hypotheses of why your friend is late -
traffic is bad and so your friend is delayed in traffic, or your friend forgot. So if your friend is generally
reliable, you may place a higher prior weight on the hypothesis that traffic was bad. While waiting, 
you overhear from someone else about an accident on the road that has caused a large traffic jam.
In light of this new information (data), you update your prior beliefs about the two hypotheses and would now
place even more weight on the traffic jam hypothesis than the "forgot" hypothesis. The updated beliefs are
your posterior beliefs about the hypotheses which is a combination of prior beliefs in the two hypotheses (prior distribution)
and information (data, or likelihood). It would be sensible to make statements such as
"I have a 75% belief that my friend is stuck in a traffic jam". Note that your friend either is
or is not stuck in traffic -- there is no probability associated with the actual state.
As more and more data are received about the size of the traffic jam, your
posterior beliefs about the two hypotheses will shift.

Similarly, you may have a prior belief about a persons age who you have never met. In this
case, the prior belief is not two discrete hypotheses, but a continuum, i.e. you can picture a "normal"-like distribution
of prior belief with a peak, say at age 50, but ranging from 40 to 60. As you get more information (data)
such as the fact that person witnessed the fall of the Berlin Wall as a young adult, you would update your
prior belief by shifting the peak of the prior age distribution and/or shrinking the range of possible ages.
This is now your posterior belief summarized by a posterior distribution. It would be sensible to make statements
such as "I have a 40% posterior belief that this person is more than 55 years of age". Note that the actual age 
of the person does NOT have a distribution -- the distribution refers to your knowledge of the age
based on a combination of prior information and actual data.
If you find the year of birth
of the person, the posterior distribution becomes extremely concentrated because you now know the age of the
person to within a year.

In an analogous fashion, a Bayesian analysis summarizes the posterior belief about a parameter using the posterior distribution.
This distribution could be discrete (e.g. two competing hypotheses) or could be continuous (e.g. a range of values summarized
by a distribution with a peak and spread.) 

We summarize the posterior distribution in this report in several ways:

* Mean of the posterior distribution the (mathematical) expectation of the belief distribution representing our point estimate 
about the parameter value.
* Standard deviation of the posterior distribution representing a measure of uncertainty about our belief in the parameter value.
* A credible interval (typically a 95% credible interval) representing uncertainty in the belief. A 95% credible interval
would contain 95% of our belief.
* Total belief that the parameter is above/below a certain value found as the portion of the posterior distribution
that is above/below the specified value. As seen below, we are interested in our posterior belief that the
slope in marginal *PIKE* was less than zero in the last five years representing our posterior belief that the 
marginal *PIKE* has declined in the last five years.

A rough correspondence exists between the results of a least-squares or likelihood model and a Bayesian model. The 
maximum likelihood estimate roughly corresponds to the mean of the posterior distribution; 
the standard error for a parameter from a likelihood fit roughly corresponds to the
standard deviation of the posterior distribution; a confidence interval from
a likelihood fit roughly corresponds to a credible interval from a Bayesian model. However,
the correspondence is not "exact" and there are fundamental differences in the technical definitions
of each measure that make them not commensurable. In particular, the total belief that a parameter is
above/below a specific value has no correspondence in the likelihood context.

## Choice between the weighted and unweighted marginal mean *PIKE*

The simple combined *PIKE* (i.e. total illegally killed elephants / total observed carcasses) 
may be a suitable estimator for the marginal mean *PIKE* if carcasses reported were proportional to
the population abundances (if you are interested in a weighted *PIKE*) or equal across sites (if you are interested
in an unweighted *PIKE*) and if all sites reported in all years.
The number of carcasses reported is not equal across sites, bears only a weak relationship
to the population abundances, and there is much missing data so the simple 
combined *PIKE* is difficult to interpret and seldom of interest.

The choice between the weighted and unweighted marginal mean *PIKE* is more complex.

The unweighted marginal mean *PIKE* should be 
thought of as an index of poaching pressure that gives each site the same weight 
regardless of the underlying population abundance.
We hope
that trends in the index are broadly informative of changes in poaching pressure at the continental scale. 

The weighted marginal mean *PIKE* may
reflect actual trends in poaching at the population level under several assumptions (e.g. that the observed *PIKE* in a MIKE site
is reflective of the poaching in the larger population; that the population abundance estimate for a MIKE site is reflective
of the actual population abundance; that all populations are monitored (i.e. every population has a MIKE site).
Consequently,
the weighted marginal mean *PIKE* looks appealing but also must make many assumptions before reflecting the poaching pressure
on a population level.

**The use of population values to compute a population-weighted PIKE is currently
experimental and being evaluated and the actual implementation may change in the future.**

Given the very weak relationship between the number of carcasses and the population abundances, the
simple combined *PIKE* will tend to track the weighted marginal mean *PIKE* closer than the unweighted marginal mean *PIKE* values.


# Continental level trends in *PIKE*

```{r echo=FALSE,warning=FALSE,message=FALSE,results="hide"}

all.fit <- fit.pike(pike, mike.pop.est,
                     seed=234234
                     )

```

The GLMM model was coded using $BUGS$ (Lunn et al, 2012), a common way to specify Bayesian 
models and run using $JAGS$ (Plummer, 2003) within $R$ (R Core Team, 2020).

The estimated *LSMeans*, unweighted and weighted marginal mean  *PIKE* are:

```{r echo=FALSE, message=FALSE, warning=FALSE, results="as.is"}
YearP.est.MM <- extract.effect(all.fit,  effect.name="^YearP.est.MM\\[", index.type="year", source="MM.p.uw")
YearP.est.MMw <- extract.effect(all.fit,  effect.name="^YearP.est.MMw\\[", index.type="year", source="MM.p.w")

temp.est.MM <- YearP.est.MM[,c("year","mean","sd")]
names(temp.est.MM) <- c("Year","MM.u","MM.u.sd")
temp.est.MMw <- YearP.est.MMw[,c("year","mean","sd")]
names(temp.est.MMw) <- c("Year","MM.w","MM.w.sd")
temp <- merge(temp.est.MM, temp.est.MMw)
temp <- temp[ order(temp$Year),]

temp[,2:5] <- round(temp[,2:5,2],2)

kable(temp, 
      caption="Estimated marginal mean PIKE from the GLMM",
      col.names=c("Year","Mean","SD","Mean","SD"),
      digits=c(0,2,2,2,2))  %>% 
      add_header_above(c(" "=1, "Unweighted "=2, "Weighted "=2)) %>%
      column_spec(column=c(1),   width="1cm") %>%
      column_spec(column=2:5,    width="1.5cm")%>%
      kable_styling("bordered",position = "center", full_width=FALSE, latex_options = "HOLD_position" )

#colnames(temp) <- c("Year","UW mean", "UW SD", "W mean", "W SD")
#library(pander)
#pander(temp)

```

The following plot compares the *PIKE* computed using the $LSMeans$ approach (the current approach), 
the unweighted (*MM.p.uw*) and weighted marginal
mean (*MM.p.w*) computed using the Bayesian model

```{r echo=FALSE, warning=FALSE, message=FALSE}
# create a plot of the year effects over time
raw.pike <- plyr::ddply(pike,"year", plyr::summarize,
                        mean=sum(NumberOfIllegalCarcasses)/sum(TotalNumberOfCarcasses))
raw.pike$Source <- "Observed"

plotdata <- plyr::rbind.fill(YearP.est.MM, 
                             YearP.est.MMw,
                             raw.pike, 
                             lsmeans.cont)

ggplot(data=plotdata, aes(x=year, y=mean, color=Source))+
  ggtitle(paste(UNRegion.select,": Estimated PIKE across time",sep=""),
          subtitle="Comparison of observed, LSMeans, unweighted and weighted marginal mean PIKE")+
  geom_point( position=position_dodge(w=0.2))+
  geom_line(position=position_dodge(w=0.2))+
  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=.2, position=position_dodge(w=0.2))+
  ylim(0,1)+
  ylab("Estimated PIKE (95% credible interval)")+xlab("Year")+
  theme(legend.position=c(0,1), legend.justification=c(0,1))
```


The new proposed unweighted marginal mean ($MM.p.uw$) tracks the previously computed $LSMeans$
fairly well except for 2009-2011. The fitted trend lines are consistently higher than the
observed *PIKE* values because the latter gives more weight to sites with more carcasses whereas the
$MM.p.uw$ give equal weight to each site regardless of number of carcasses (or underlying
population abundance). The weighted marginal mean ($MM.p.w$) tracks the observed *PIKE* closely after 2010 because
sites with large number of carcasses (and larger elephant populations) will dominate both in a similar way.


```{r include=FALSE,echo=FALSE, warning=FALSE, message=FALSE}
# Estimate the posterior probability that the trend in the last five years is negative.
# Extract the posterior sample for the margina mean (unweighted) for the last five years
yearP.eff.post.long <- extract.posterior(all.fit, "^YearP.est.MM\\[", index.type="year" )

# select the last five years
last.five.years <- sort(unique(pike$year),decreasing=TRUE)[1:5]
yearP.eff.post.long <- yearP.eff.post.long[ yearP.eff.post.long$year %in% last.five.years, ]

# compute the slope in pike in the last five years for each sample from the posterior
yearP.eff.slope <- plyr::ddply(yearP.eff.post.long, "sim", function(x){
    fit <- lm( value ~ year, data=x)
    slope <- coef(fit)[2]
    slope.neg <- slope < 0
    data.frame(slope=slope, slope.neg=slope.neg)
})
#head(yearP.eff.slope)
```


```{r echo=FALSE, warning=FALSE, message=FALSE}
post.belief.slope.negative <- formatC(mean(yearP.eff.slope$slope.neg), digits=2, format='f')


slope.post <- ggplot(data=yearP.eff.slope, aes(x=slope, y=..density..))+
  ggtitle("Posterior distribution of slope in fitted yearly PIKE in last five years",
          subtitle="Based on unweighted marginal mean of back transformed to [0,1] scale - MM.p.uw")+
  geom_histogram( alpha=0.2)+
  geom_vline(xintercept=0)+
  annotate("text", label=paste("Post belief that slope is < 0 is ", post.belief.slope.negative,sep=""),
            x=Inf, y=Inf, hjust=1, vjust=1)+
  xlab("Slope in estimated PIKE in last five year")+
  ylab("Density")
#slope.post
```

Once the sample from the posterior is available, it is relatively easy to estimate the posterior belief that the
trend is negative in the last five years. This is done by estimating the slope in the last five years
for each sample from the posterior and then the posterior belief that the trend is negative is the proportion 
of fitted slopes that are less than zero. The posterior belief that the slope in PIKE
is negative in the last five years is `r post.belief.slope.negative`, i.e. we have a high posterior
belief that the trend in PIKE in the last five years is negative (i.e., is declining).



```{r echo=FALSE, warning=FALSE, messge=FALSE}
# cleanup
rm(slope.post, post.belief.slope.negative,last.five.years )
rm(yearP.eff.slope, yearP.eff.post.long, plotdata)
###############################################################################################3
##############################################################################################3
#############################################################################################3
```




# Subregional trends in *PIKE*

The above analyses were repeated at the sub-regional level. Only the data from each sub-region was used in
each analysis, i.e., completely separate analyses were performed for each sub-region. 

A more complex model where
the four sub-regions were modelled as part of a larger model could be attempted.
This larger model would be useful in cases where a sub-region has sparse data but has 
a consistent relationship in *PIKE* trends with other sub-regions. This is similar
to the case where sparse data for a particular MIKE site is improved if there is a
consistent relationship between the *PIKE* in several MIKE sites. In these
cases, the consistency in the relationship allows information from other
sub-regions to improve the estimates of trend in the sub-region with sparse data.

This more complex model is examined in the technical document, but because there does not appear to be 
a great deal of similarity in trends among sub-regions,  we do not expect
this more complex model to provide much of an improvement over individual
models for each sub-region.

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
# set up individual seeds for subregion fit
set.seed(2343243)
subregion.seed <- data.frame(SubregionName=unique(pike$SubregionName))
subregion.seed$seed <- round(runif(nrow(subregion.seed),min=1, max=1000000000))

reg.fit <- plyr::llply( unique(pike$SubregionName), function(subregion, pike, mike.pop.est, subregion.seed){
   # restrict the PIKE and pop est and seed to this subregion
  pike           <- pike           [ pike          $SubregionName == subregion,] 
  mike.pop.est   <- mike.pop.est   [ mike.pop.est  $SubregionName == subregion,]
  subregion.seed <- subregion.seed [ subregion.seed$SubregionName == subregion,]
  fit <- fit.pike(pike=pike, mike.pop.est=mike.pop.est, seed=subregion.seed$seed)
  fit
}, pike=pike, mike.pop.est=mike.pop.est, subregion.seed=subregion.seed)

# Get the various sources to plot
reg.raw.pike <- plyr::ddply(pike,c("SubregionName","year"), plyr::summarize,
                        mean=sum(NumberOfIllegalCarcasses)/sum(TotalNumberOfCarcasses))
reg.raw.pike$Source <- "Observed"

# marginal mean on [0,1] scale
reg.YearP.eff <- plyr::ldply(reg.fit, function(x){
   effect    <- extract.effect(x,  effect.name="^YearP.eff\\[",    index.type="year", source="MM.logit")
   effect$SubregionName = x$pike$SubregionName[1]
   effect
})

# The unweighted marginal mean of the $PIKE$ computed on the [0,1] scale is:
reg.YearP.est.MM <- plyr::ldply(reg.fit, function(x){
  effect <- extract.effect(x,  effect.name="^YearP.est.MM\\[", index.type="year", source="MM.p.uw")
   effect$SubregionName = x$pike$SubregionName[1]
   effect
})

#F the weighted marginal mean of the $PIKE$ computed on the [0,1] scale is:
reg.YearP.est.MMw <- plyr::ldply(reg.fit, function(x){
   effect <- extract.effect(x,  effect.name="^YearP.est.MMw\\[", index.type="year", source="MM.p.w")
   effect$SubregionName = x$pike$SubregionName[1]
   effect
})

```

The following plots compare the mean *PIKE* computed using the $LSMeans$ method,
the unweighted marginal mean *PIKE*  (*MM.p.uw*) and the weighted marginal mean *PIKE* (*MM.p.w*).

```{r echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
reg.plotdata <- plyr::rbind.fill(#reg.YearP.eff,  # don't use the back transformed mean of logits 
    reg.YearP.est.MM, 
    reg.YearP.est.MMw,
    reg.raw.pike, 
    lsmeans.reg)

  reg.plot <- ggplot(data=reg.plotdata[], 
                     aes(x=year, y=mean, color=Source))+
     ggtitle(paste(UNRegion.select,": Estimated subregional PIKE trends",sep=""))+
     geom_point( position=position_dodge(w=0.2))+
     geom_line(position=position_dodge(w=0.2))+
     geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=.2, position=position_dodge(w=0.2))+
     coord_cartesian(ylim=c(0,1))+
     ylab("PIKE\n (95% credible interval)")+xlab("Year")+
     theme(legend.position=c(.5,.5), legend.justification=c(1,0), 
           legend.background=element_blank(),legend.key=element_blank(), 
           legend.spacing.y=unit(-.8,"mm"), legend.text=element_text(size=10), legend.key.size = unit(1, 'lines'))+
     facet_wrap(~SubregionName, ncol=2, nrow=2)
  reg.plot

```

The estimates at each subregion have wider confidence intervals because the amount of data is smaller in each subregion
compared to the continental results.

It is hard to make a general statement comparing the estimates computed using $LSMeans$ and the
unweighted marginal means (*MM.p.uw*) because of the wide confidence intervals, but the respective trend
lines mostly track each other. The apparent consistent difference seen for Eastern Africa and
Southern Africa are artefacts of how data are aggregated to the sub-regional level 
in the *LSMeans* and Bayesian GLMM models (see Appendix 3 of the technical document for details).

In Central Africa, the weighted marginal mean *PIKE* is consistently higher 
than the unweighted marginal mean *PIKE* but the 
credible intervals overlap); 
in Eastern and West Africa, they track each other closely; 
while in Southern Africa, the unweighted marginal mean *PIKE* is consistently higher
than the weighted mean *PIKE*. This "inconsistency" is a function of 
the relationship between the number of carcasses examined and the 
population abundance examined earlier in this document.

But regardless if the weighted or unweighted marginal mean *PIKE* are examined, trends are similar. 

**The use of population values to compute a population-weighted PIKE is currently
experimental and being evaluated and the actual implementation may change in the future.**

It is interesting to compare the regional trends with the continental trends

```{r echo=FALSE, warning=FALSE, message=FALSE}
cont.plotdata <-  plyr::rbind.fill( 
                             YearP.est.MM, 
                             YearP.est.MMw)
cont.plotdata$SubregionName = " Continental"
cont.plotdata$size          =2
cont.plotdata$color         ="black"

reg.plotdata <- plyr::rbind.fill(#reg.YearP.eff,  # don't use the back transformed mean of logits 
    reg.YearP.est.MM, 
    reg.YearP.est.MMw)

ggplot(data=reg.plotdata, aes(x=year, y=mean, color=SubregionName))+
     ggtitle(paste(UNRegion.select,": Estimated subregional PIKE across time",sep=""),
             subtitle="Continental trend shown in black")+
     geom_point( position=position_dodge(w=0.2))+
     geom_line(position=position_dodge(w=0.2))+
     geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=.2, position=position_dodge(w=0.2))+
     coord_cartesian(ylim=c(0,1))+
     ylab("Estimated PIKE (95% credible interval)")+xlab("Year")+
     #theme(legend.position=c(0,1), legend.justification=c(0,1), legend.background=element_blank(),legend.key=element_blank())+
     facet_wrap(~Source, ncol=1, nrow=2)+
     geom_line(data=cont.plotdata, aes(x=year, y=mean), color="black", size=2)
```

We see that *PIKE* in Southern Africa is consistently lower than the continental
*PIKE*, while *PIKE* in Central Africa is consistently higher. 

```{r echo=FALSE, warning=FALSE, messge=FALSE}
# cleanup
rm(reg.plotdata, lsmeans.reg, reg.raw.pike, reg.YearP.eff, reg.YearP.est.MM, reg.YearP.est.MMw)
###############################################################################################3
##############################################################################################3
#############################################################################################3
```



# Model assessment

Whenever a new method is used, it is important to determine if the model fits the data well.
We performed detailed model assessments (see the technical document) and found no evidence
of problems except for a residual spatial autocorrelation in the estimated site effects.

## Observed vs. predicted *PIKE*

A plot of observed *PIKE* in each year.site vs. the predicted *PIKE* is:

```{r echo=FALSE, message=FALSE, warning=FALSE}
est.pike <- extract.effect(all.fit,  effect.name="^Year.SiteP.est\\[", index.type="year.site", source="Expected")
pike$obs.pike <- pike$NumberOfIllegalCarcasses/ pike$TotalNumberOfCarcasses

#quantile(pike$TotalNumberOfCarcasses, prob=seq(.1,.9,.1))

plotdata <- merge(pike, est.pike[ ,c("mean","MIKEsiteID","year")], all.x=TRUE)
plotdata$n.carcass <- car::recode(plotdata$TotalNumberOfCarcasses,
                              " lo:10='00-10'; 10:25='10-25'; 25:hi='25+'     ")
ggplot(data=plotdata, aes(x=mean, y=obs.pike, size=n.carcass, color=n.carcass))+
   ggtitle(paste(UNRegion.select,": Observed vs. Predicted PIKE"))+
   geom_point(shape=1)+
   geom_abline(intercept=0, slope=1)+
   scale_color_discrete(name="Number\nof\ncarcasses")+
   scale_size_discrete (name="Number\nof\ncarcasses")+
   #scale_colour_brewer(direction=-1)+
   xlab("Mean predicted PIKE from Bayesian model")+
   ylab("Observed PIKE")

```

The fit is generally very good. 
For site-years where the number of carcasses was very small ($<10$) and the observed
*PIKE* was 0 or 1, the estimated *PIKE* is pulled towards the yearly average for that year.
For site-years with large number of carcasses ($>25$) the estimated *PIKE* matches 
closely with the observed *PIKE*. For site-years with intermediate number of
carcasses, the estimates are shrunk slightly towards the mean for that year.

This can also be seen in the plots of observed and fitted *PIKE* for the individual MIKE sites:

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Make a plot of trend over time

plotdata <- merge(pike, est.pike, all.y=TRUE)
plotdata$n.carcass <- car::recode(plotdata$TotalNumberOfCarcasses,
                              " lo:10='00-10'; 10:25='10-25'; 25:hi='25+'     ")

YearP.eff.MM <- as.data.frame(extract.effect(all.fit,  effect.name="^YearP.est.MM\\[",    index.type="year", source="All sites"))

plotdata$X2.5.  <- pmax(0, pmin(1, plotdata$X2.5.))
plotdata$X97.5. <- pmax(0, pmin(1, plotdata$X97.5.))

plyr::l_ply(1:ceiling(length(unique(plotdata$MIKEsiteID))/16), function(page){
  myplot <- ggplot(data=plotdata[!is.na(plotdata$obs.pike),], aes(x=year, y=obs.pike))+
   ggtitle(paste(UNRegion.select,": Observed and predicted PIKE for individual MIKE sites"))+
   geom_point(aes(size=n.carcass, color=n.carcass), shape=1)+
   geom_line(data=plotdata,aes(y=mean), color="blue")+
   geom_ribbon(data=plotdata, aes(ymin=X2.5., ymax=X97.5.), alpha=0.2, fill="blue", linetype=0)+
   geom_line(data=YearP.eff.MM, aes(y=mean), color="black", size=.5, linetype="dashed", alpha=0.5)+
   #scale_colour_brewer(direction=-1)+
   xlab(paste0("Year\nDashed line is unweighted marginal mean PIKE at continental level",
               "\nBlue and shading is predicted PIKE at site level with 95% credible interval"))+
   ylab("PIKE")+ylim(0,1)+
   scale_color_discrete(name="Number\nof\ncarcasses")+
   scale_size_discrete (name="Number\nof\ncarcasses")+
   facet_wrap_paginate(~MIKEsiteID, ncol=4, nrow=4, scales="free_y", page=page)
  plot(myplot)
})
```

There are several interesting patterns that illustrate the features of the model.

* Site *CHO*. This site reports nearly every year with a large number of carcasses (large blue circles). The estimated
yearly site-level *PIKE* closely follows the observed data (as expected).

* Site *BBK*. In some years, this site reports a large number of carcasses and the estimated yearly site-level *PIKE*
for these years matches the observed *PIKE*. In some years, the observed *PIKE* based on large number of carcasses is
above the continental marginal mean (e.g. 2003) and in some some years it is below the continental marginal mean (e.g. 2006).
On average, this site tracks the continental trend. So in years where this site only reports on a few carcasses (small red circles) and
the observed *PIKE* is mostly 0 or 1,
the estimated *PIKE* is close to the continental trend.
For example, with a small number of carcasses examined, a value of 2 illegally killed elephants from 2 carcasses
examined (observed *PIKE* of 1) is consistent with an estimated *PIKE* closer to the continental marginal *PIKE*. 
Notice that in years with a small number of carcasses reported, the credible interval for the estimated site-level
*PIKE* is very wide.

* Site *AKG*. This site mostly has smallish sample sizes, but the observed *PIKE* is consistently close to 0.
The estimated site-level *PIKE* is then also close to 0 in years with no reports, but notice the wide 
credible intervals.

* Site *KFE*. This is a new MIKE site with data only starting in 2018. In this year, the observed *PIKE* 
was slightly higher than the continental marginal mean *PIKE*, and so the estimated site-level *PIKE* for
earlier years is slightly above the continental trend, but notice the wide credible intervals.

In summary, in years with many carcasses reported, the estimated site-year *PIKE* will closely match the observed 
site-year *PIKE*.
In years with few carcasses reported, the estimated
site-year *PIKE* will be pulled towards the continental trend after accounting for the observed
relationship between this sites *PIKE* and the continental trend.

## Spatial correlation in site effects

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Extract the site effects and merge with site centroids
site.eff <- extract.effect(all.fit, effect.name="^Site.eff", index.type="site")
site.eff <- merge(site.eff, mike.centers[, c("MIKEsiteID","lon","lat")], all.x=TRUE)
site.eff$posneg <- ifelse(site.eff$mean <0, "Below\nmean", "Above\nmean")

site.effects.spatial.plot <- ggplot(data=site.eff, aes(x=lon, y=lat, size=abs(mean), color=posneg))+
  ggtitle(paste(UNRegion.select,": Estimated site effects", sep=""))+
  geom_point(alpha=0.6)+
  theme(legend.position=c(0,0), legend.justification=c(0,0), legend.direction = "vertical", legend.box = "horizontal")+
  scale_size_continuous(name="Absolute \neffect\nsize")+
  scale_color_discrete (name="Direction\neffect\nsize")+
  xlab("Longitude")+ylab("Latitude")
#site.effects.spatial.plot

site.effects.spatial.plot2 <- base.map2 +
  geom_point(data=site.eff, aes(x=lon, y=lat, size=abs(mean), shape=posneg, color=posneg))+
  ggtitle(paste(UNRegion.select,": Estimated site effects", sep=""))+
  geom_point(alpha=0.6)+
  #theme(legend.position=c(0,0), legend.justification=c(0,0), legend.direction = "vertical", legend.box = "horizontal")+
  scale_size_continuous(name="Absolute \neffect\nsize")+
  scale_color_manual (name="Direction\nsite\neffect", values = c("red","green"))+
  scale_shape_discrete (name="Direction\nsite\neffect")+
  xlab("Longitude")+ylab("Latitude")+xlim(-20,50)
```

The (random) site effects have been modelled as independent random effects without explicitly
accounting for the spatial structure of the data. 
However, we find that sites that are close geographically have similar site effects.


```{r echo=FALSE, warning=FALSE, message=FALSE}
#site.effects.spatial.plot
site.effects.spatial.plot2
```

Sites that have  *PIKE* consistently above the continental average are labelled as *Above the mean*; 
sites that have *PIKE*
consistently below the continental average are labelled as *Below the mean*.

We notice that sites that are close geographically tend to have similar site effects 
(size of dot) and in the same direction (above or below the mean, color of dots). This implies
there is a spatial correlation among the site effects that has not been directly accounted
for in the analysis.

The current analysis is still valid, but inefficient because it has not used the
spatial correlation to improve inference. If spatial autocorrelation is explicitly modelled, then
information is shared among sites that are geographically close, i.e., if the *PIKE* increases
in one site, then spatial autocorrelation would imply that it would tend to also increase in a nearby site.
Of course, if the sites are in different countries with different levels of enforcement or other 
covariates that impact *PIKE*, an explicit spatial autocorrelation could introduce a spurious relationship between
the *PIKE* in the two sites unless these other factors (law enforcement etc.) are also modelled. 
The explicit spatial autocorrelation models rapidly become more complex to account for these features.

Because the current analysis treats all sites as independent (rather than spatially correlated), the uncertainty 
in the overall yearly *PIKE* is slightly smaller than from a model with explicity spatial autocorrelation because the
effective number of sites used in computing the overall yearly *PIKE* is smaller when autocorrelation is explicitly modelled.
This in turn, implies that the uncertainty of a trend (e.g. trend in the last five years) in the currently model may 
be slightly understated as well and the posterior belief in a trend will be higher in the current model compared to the
model with an explicity spatial autocorrelation. We believe such effects are minor given the spase data at many sites,
the large amount of missing site.years and the potential breaking of spatial autocorrelation across country borders. 

A potential improvement to the current analysis may be to add another level of random effects (country effects) so that
points from the same country that have related site effects then experience a common
country effect. This model is currently under investigation.

## Sensitivity analysis

We also examined in the technical document, the sensitivity of the new model to 
adding new MIKE sites, to dropping MIKE sites (perhaps due to local extirpation), 
increasing monitoring effort at MIKE sites, and the impact of the prior beliefs as part of the Bayesian model.


Adding a new MIKE site may have impacts on previous estimated *PIKE* values. 
Presumably the MIKE site had elephants and potential
illegal killings prior to being added to the programme. 
The current model will impute the
*PIKE* for this site in the years before the site was added based on the yearly trend in *PIKE* in other sites
and the relationship between the new site's *PIKE* and other sites' *PIKE*. For example,
if the new MIKE site had a *PIKE* value that was 10 percentage points above the mean *PIKE*, then 
a *PIKE* will be imputed for this site for all past years that is also 10 percentage points above the yearly mean.

The imputation will be very coarse in the first few years after a MIKE site is added
until enough yearly *PIKE* data has been collected to estimate reasonably well the
relationship between this site's *PIKE* and the overall mean (i.e., several years will
be needed to estimate the (random) $site$ effect well).

Fortunately, the (unweighted) marginal mean *PIKE* is currently computed using over 50 MIKE sites
and so adding a new MIKE site is expected to only have minimal impact on the yearly mean *PIKE* values
unless the new site has an extreme *PIKE*. The weighted marginal mean *PIKE* could be influenced
more if the new MIKE site represented a large elephant population that was not previously monitored.

In most cases, the effect of dropping individual sites is slight 
except  when *PIKE* is computed by weighting by population abundance.
In these cases, the yearly mean *PIKE* values can change substantially when a site with a large
underlying population is dropped.
This influence could push the year mean *PIKE* up or down depending if the particular site has a site-specific
*PIKE* that was larger or smaller than the overall yearly mean *PIKE*.

Increasing monitoring effort at MIKE sites will shrink the width of credible intervals, but missing site-years
is the limiting factor in determining the uncertainty of the marginal mean *PIKE* because imputation must be
performed for sites with missing values and increasing the sample size at other MIKE sites has little impact on the
uncertainty in the imputation.

The data is rich enough (more than 20 years with more than 50 sites) so there prior belief distributions
have minimal impact on the results.

We therefore conclude that the proposed
GLMM model is robust.

# Conclusions

The GLMM method is an improvement over the current method (the *LSMeans* method) by accounting for the
binomial structure of the data, the interrelationship between *PIKE* among sites, and for local variations
in *PIKE* at the site level over time. The results from the proposed model are similar to the current results
and conclusions about trends in *PIKE* are the same under both models. The GLMM model can also be easily 
extended to finding weighted marginal mean *PIKE* (e.g. weighted by population abundances)
and posterior beliefs about features of the data (e.g. trend
in the last five years) that are impractical to obtain from the *LSMeans* approach. The GLMM model is also
robust to changes in the data (i.e. not sensitive to data changes). 


# References
Burn, R.W., Underwood, F.M., Blanc J. (2011).
Global Trends and Factors Associated with the Illegal Killing of Elephants:
A Hierarchical Bayesian Analysis of Carcass Encounter Data. PLoS ONE 6(9): e24165.

Lunn, D., Jackson, C., Best, N., Thomas, A. and Spiegelhalter, D. (2012). 
The BUGS Book  A practical introduction to Bayesian Analysis. 
Chapman and Hall/CRC Press.

Plummer, M. (2003). 
JAGS: A program for analysis of Bayesian graphical models using Gibbs sampling. 
Proceedings of the 3rd International Workshop on Distributed Statistical Computing (DSC 2003), March 2022, Vienna, Austria. ISSN 1609-395X.

R Core Team (2020). 
R: A language and environment for statistical computing. 
R Foundation for Statistical Computing, Vienna, Austria.

